---
title: "HW4"
author: "Louise Lai"
date: "August 13, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(stringr)
library(dplyr)
library(magrittr)
```

# Edge Case Implementation Choices:
- When a number/roman numeral is in the last position, delete it
- When there's only one name (i.e. no last name), use it as both the first and last name

## Handles:
- All explicit specifications from homework
- 1st, 2nd, 3rd, 4th, 1ST, 2ND, 3RD, 4TH, 200000th etc.
- NAMES IN ALL CAPS

## Fails when:
- When last name is similar to a roman numeral, e.g. "Louise Ying She LAI". The program intprets the last name as a roman numeral and deletes it.
- When the name contains a number, but the number is not in the last position e.g. "Victoria 2nd Smith". The program does not properly locate number. Instead, it just deletes the last word.

```{r}

nameChanger <- function(name){
  
  nameSplit <-  unlist(strsplit(name, "\\s+")) # split with at least one space
  # 1) Non-conforming inputs
  
   # case: name too long
  if(length(nameSplit) > 2){
    message("input non-conforming! (name > 2)")
  } 
  
   # case: numbers and/or roman numerals
  edgeCaseRegex <- "(^\\d+[(st)|(nd)|(rd)|(th)|(ST)|(ND)|(RD)|(TH)]*$)|(XC|XL|L?X{0,3})(IX|IV|V?I{1,3})$"
  if(sum(!is.na(str_extract(nameSplit, edgeCaseRegex))) > 0){
    # delete the word. Only works if number is at the end, and there is one. 
    nameSplit <- nameSplit[-length(nameSplit)]
    message("input non-conforming! (numbers/roman numerals)")
  }

  # 2) Charts
  firstChart <- unlist(strsplit("Stinky Lumpy Buttercup Gidget Crusty Greasy Fluffy Cheeseball Chim.Chim Poopsie Flunky Booger Pinky Zippy Goober Doofus Slimy Loopy Snotty Falafel Dorky Squeezeit Oprah Skipper Dinky Zsa-Zsa", " "))
  secondChart <- unlist(strsplit("Diaper Toilet Giggle Bubble Girdle Barf Lizard Waffle Cootie Monkey Potty Liver Banana Rhino Burger Hamster Toad Gizzard Pizza Gerbil Chicken Pickle Chuckle Tofu Gorilla Stinker", " "))
  thirdChart <- unlist(strsplit("head mouth face nose tush breath pants shorts lips honker butt brain tushie chunks hiney biscuits toes buns fanny sniffer spinkles kisser squirt humperdinck brains juice", " "))
  
  # 3) Create Dictionary
  
  # get names
  firstName <- tolower(nameSplit[1])
  lastName <- tolower(nameSplit[length(nameSplit)])

  # get letters as values
  firstLetter <- str_match(firstName, "[a-z]")
  secondLetter <- str_match(lastName, "[a-z]")
  thirdLetter <- str_match(stringi::stri_reverse(lastName), "[a-z]")
  
  # this error should never happen, but just in case truly weird input occurs
  if(is.na(firstLetter) | is.na(secondLetter) | is.na(thirdLetter)){
    message("input non-conforming! Super weird! Aargh!\n") 
  }
  
  # assign keys
  names(firstChart) <- letters[1:26]
  names(secondChart) <- letters[1:26]
  names(thirdChart) <- letters[1:26]

  # 4) Match and Return
  print(paste(name, " -> ", firstName, lastName)) # previous 
  print(paste(firstChart[firstLetter], paste(secondChart[secondLetter], thirdChart[thirdLetter], sep=""))) # new
  print("- - - - - - - -") # purely aesthetic, for output
}

names <- c("Louise Lai", "Louise Ying She Lai", "Paul Intrevado", "David Uminsky", "Terence Parr", "Jeff Hamrick",
           "paul intrevado", "Intrevado, Paul", "Intrevad0 Paul", "Queen Elizabeth II", 
           "Queen Elizabeth 2nd", "Queen Elizabeth 20000TH", "John Paul Euclid Rumpel", "britishDudeThatSitsInTheBackOfTheClass")

sapply(c("Louise Lai", "Louise Ying She Lai", "Paul Intrevado", "David Uminsky", "Terence Parr", "Jeff Hamrick",
           "paul intrevado", "Intrevado, Paul", "Intrevad0 Paul", "Queen Elizabeth II", 
           "Queen Elizabeth 2nd", "Queen Elizabeth 20000TH", "John Paul Euclid Rumpel", "britishDudeThatSitsInTheBackOfTheClass"), nameChanger)
```


