---
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Q2

Trying to load all 28 CSV documents kept crashing my (rather frail) Lenovo computer - I apologize for not getting this to work, my Macbook Pro is on the way!

\newpage
# Q3
```{r, echo=F, eval=T}
setwd("/home/louiselai88gmail/Desktop/programming/eda/homework/hw2")
heartAttack_1 <- read.csv("./data/heart-attack.csv", stringsAsFactors = T)
heartAttack_2 <- read.csv("./data/heart-attack.csv", stringsAsFactors = F)
heartAttack_3 <- read.csv("./data/heart-attack.csv")

str(heartAttack_1)
str(heartAttack_2)
str(heartAttack_3)
```

```{r}
levels(heartAttack_1$work_type)

# min(heartAttack_1$work_type) # not ordered

barplot(table(heartAttack_1$work_type))

heartAttack_1$work_type <- ordered(heartAttack_1$work_type)

barplot(table(heartAttack_1$work_type))

barplot(table(heartAttack_1$gender)) # there are spelling differences that create 5 factors instead of 3

heartAttack_1$gender[heartAttack_1$gender == "Male"] <- "male"

barplot(table(heartAttack_1$gender)) # the factor still exists, even though there are 0 observations in it

heartAttack_1$gender[heartAttack_1$gender == "Female"] <- "female"

table(droplevels(heartAttack_1)$gender)

```

\newpage
# Q4
Create a vector of 10,000 random variates $\sim \mathcal{U}\{10^{-15},10^{5}\}$. These numbers represent wavelengths (in meters) of photons hitting the Hubble telescope. 
```{r}
wavelengthsNums <- exp(runif(10000,log(1e-15),log(1e5)))
wavelengths <- exp(runif(10000,log(1e-15),log(1e5)))
```

- Convert this vector into an ordinal factor with levels as the type of wave. Levels should be ordered in terms of increasing wavelength. 
```{r}
library("ggplot2")
gamma <- 1e-11
xray <- 10e-8
uv <- 4 * 10e-7
visible <- 7.5 * 10e-7
infrared <- 10e-6
microwave <- 10e-3
radio <- 10e3

frequencies <- c(gamma, xray, uv, visible, infrared, microwave, radio)

wavelengths <- cut(wavelengths,
                   breaks=c(-Inf, gamma, xray, uv, visible, infrared, microwave, radio, Inf),
                   ordered_result = TRUE,
                   right=TRUE,
                   labels=c("gamma", "xray", "uv", "visible", "infrared", "microwave", "radio", "beyondRadio")
                   )
```

 - Create a boxplot for each factor level.
```{r}
ggplot(data.frame(wavelengths, wavelengthsNums), aes(x=wavelengths, y=log(wavelengthsNums), fill=wavelengths))+
  geom_boxplot() +
  labs(title="Plot of Wavelengths and Log Wavelength ")
```
 
 - How many photons can you see with the the naked eye?
```{r}
10000 * ((visible-uv)/(10e5-10e-15))
```
 
 