---
title: "Traffic Exploration"
author: "Louise Lai"
date: "August 4, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(readr)
library(lubridate)
library(magrittr)
library(forcats)
library(ggplot2)
library(scales)
library(viridis)
library(ggthemes)
```

```{r}
# load files
setwd("~/Desktop/programming/eda/homework/hw3")
original <- as.data.frame(read.csv("./Officer_Traffic_Stops.csv", na.strings = c("", " ", "  ", "NA"))) 

```

# Data Cleaning
```{r}

df <- original

##### Data Cleaning

returnCleanData <- function(){
  
  # count NAs
  anyNA(df) # TRUE
  colnames(df)[apply(df, 2, anyNA)] # returns colnames that have NAs 
  
  ## NAs
  df$Officer_Race <- as.character(df$Officer_Race)
  df %<>% mutate(Officer_Race = replace(Officer_Race, is.na(Officer_Race), "Missing")) # replace
  df$Officer_Race <- as.factor(df$Officer_Race)
  
  # unsure what to do with CMPD_Division NAs yet!
  
  ## Dates
  df$Month_of_Stop <- parse_date(df$Month_of_Stop, "%Y/%m")
  
  ## Logicals
  levels(df$Was_a_Search_Conducted) <- c(FALSE, TRUE) # set 'No' to FALSE, 'Yes' to TRUE
  df$Was_a_Search_Conducted <- as.logical(df$Was_a_Search_Conducted)
  
  ## Order Factors
  levels(df$Result_of_Stop) # come back~ <<
  
  # Change White/Hispanic 
  df %>% 
    filter(Driver_Ethnicity == "Hispanic") %>% 
    nrow() # There are 6,623 drivers that are White and Hispanic. 7,578 total == 955 unidentified other Hispanic. Small enough to disregard.
  
  df$Driver_Race <- as.character(df$Driver_Race)
  
  df %<>%
    mutate(Driver_Race = replace(Driver_Race, Driver_Ethnicity == "Hispanic", "Hispanic")) # replace Hispanic (ethnicity) to just Hispanic (race)
  
  # Filter truly unnecessary columns
  df %<>% 
    select(-13, -14, -15, -16, -17)
  
  return(df)
}

df <- returnCleanData()

```

\newpage
# Exploring Months
```{r}
# Frequency of Stops by Month
monthStops <- df %>% 
  group_by(Month_of_Stop) %>% 
  summarize(Freq=n())

df %>% 
  mutate(season = case_when(Month_of_Stop == as.Date("2016-06-01") | Month_of_Stop == as.Date("2016-07-01") | Month_of_Stop == as.Date("2016-08-01") ~ "Summer",
                            Month_of_Stop == as.Date("2016-03-01") | Month_of_Stop == as.Date("2016-04-01") | Month_of_Stop == as.Date("2016-05-01") ~ "Spring",
                            Month_of_Stop == as.Date("2016-12-01") | Month_of_Stop == as.Date("2016-01-01") | Month_of_Stop == as.Date("2016-02-01") ~ "Winter",
                            Month_of_Stop == as.Date("2016-09-01") | Month_of_Stop == as.Date("2016-10-01") | Month_of_Stop == as.Date("2016-11-01") ~ "Autumn")) %>% 
  group_by(Month_of_Stop) %>% 
  ggplot() +
    geom_bar(aes(Month_of_Stop, fill=factor(season))) +
    scale_fill_brewer(palette = "Spectral", name="Season") +
    xlab("Month of Stop") +
    ylab("Number of Stops Recorded") +
    labs(title= "Number of Stops per Month",
         subtitle="Subset by Season") +
    theme_economist() 

```
The Spring and Summer months have the highest levels of stops recorded, and it
appears that Autumn and Winter are, on average, the lowest months. 

This could be becuase people are more likely to drive around when the weather is nice,
such as in the Spring or the end of Winter. Conversely, at the start of Winter, there
are far fewer stops made.

\newpage
# Exploring Years of Service
```{r}
# years of service ~ average stops
serviceStops <- df %>% # creating bucket in 5 yr increments
  mutate(yearBucket = case_when(Officer_Years_of_Service < 5 ~ "<5", 
                                (Officer_Years_of_Service >= 5) & (Officer_Years_of_Service < 10) ~ "5 - 10",
                                (Officer_Years_of_Service >= 10) & (Officer_Years_of_Service < 15) ~ "10 - 15",
                                (Officer_Years_of_Service >= 15) & (Officer_Years_of_Service < 20) ~ "15 - 20",
                                (Officer_Years_of_Service >= 20) & (Officer_Years_of_Service < 25) ~ "20 - 25",
                                (Officer_Years_of_Service >= 25) & (Officer_Years_of_Service <= 30) ~ "25 - 30",
                                Officer_Years_of_Service > 30 ~ ">30")) %>% 
  select(Officer_Gender, yearBucket)

# order year buckets
serviceStops$yearBucket <- ordered(serviceStops$yearBucket, levels= c("<5", "5 - 10", "10 - 15", "15 - 20", "20 - 25", "25 - 30", ">30"))

# avg stops
serviceStops %>% 
  group_by(Officer_Gender, yearBucket) %>% 
  summarize(stops = n()) %>% 
  ggplot() +
    geom_line(aes(x=yearBucket, y=stops, group=Officer_Gender, color=Officer_Gender), size=.7) +
    geom_point(aes(x=yearBucket, y=stops, color=Officer_Gender, size=stops), shape=21, fill="white") +
    xlab("Officer's Years of Service") +
    ylab("Frequency of Stops") +
    scale_color_discrete(name = "Officer Gender") +
    #scale_color_manual(labels = c("Nope", "Yes, Busted!"), values = c("grey", "red3"), name="Was legal action taken?") +
    scale_size_continuous(name="Stops") +
    scale_y_continuous(labels=comma) +
    labs(title="Frequency of Stops", subtitle="By Gender and Officer Years of Service") +
    theme_economist() +
    theme(text=element_text(family="Helvetica"))

```
We can clearly see that the bulk of stops are made by younger officers, and the least
by senior officers. This could be due to the enthusiasm of younger officers, or that
senior officers are doing more important work. 

The _average_ number of stops made would be an ideal statistic; however, though seemingly 
straightforward, there is no actual way to derive this information from the given dataset 
because we were only given the number of stops that were made. In order to get an average,
we would need to count the total number of officers - no only officers that made a stop - in a particular year bucket and take that as our denominator to get the average.

\newpage
# Exploring Legal Action

Not every stop leads to an arrest. Some stops can result in a mere verbal warning, with the most severe being an arrest on
the spot. In this case study, we definte a legal action as receiving a citation or getting arrested. 

These fields are found in the `Result_of_Stop` column.

```{r, message=F, warning=F}
# Exploring legal action by race, age and gender
ageGender <- df %>% 
  filter(!Driver_Race == "Other/Unknown" & !Driver_Race == "Native American") %>% 
  mutate(legal_action_taken = Result_of_Stop == "Citation Issued" | Result_of_Stop == "Arrest") %>% 
  select(Driver_Age, Driver_Gender, Driver_Race, legal_action_taken) 

ageGender %>% 
  ggplot() +
  geom_point(aes(x=Driver_Race, y=Driver_Age, color=legal_action_taken, shape=Driver_Gender), alpha=0.5, position="jitter") + 
  scale_color_manual(labels = c("Nope", "Yes, Busted!"), values = c("grey", "red3"), name="Was legal action taken?") +
  scale_shape_manual(values=c(1, 2)) +
  scale_shape(name = "Driver Gender") +
  xlab("Driver Race") + 
  ylab("Driver Age") +
  labs(title="Was Legal action taken?", subtitle= "Based on Age, Race and Gender") +
  theme_few()
```

On a whole, we see that white drivers 

\newpage
# Exploring Relationships Between Columns
```{r, warning=F}
# Using alluvia, which is great for categoricla data, which we have a lot of
library(ggalluvial)

likelyAlluvia <- df %>% 
  filter(!Driver_Race=="Other/Unknown" & !Driver_Race=="Native American") %>% 
  mutate(legal_action_taken = Result_of_Stop == "Citation Issued" | Result_of_Stop == "Arrest") %>% 
  group_by(legal_action_taken, Driver_Gender, Driver_Race) %>% 
  summarise(Frequency=n())

is_alluvia_form(likelyAlluvia, axes=1:3, silent=TRUE)

likelyAlluvia %>% 
  ggplot(aes(y = Frequency, axis2 = Driver_Gender, axis1=Driver_Race)) +
  geom_alluvium(aes(fill = legal_action_taken), width = 1/12) +
  geom_stratum(width = 1/12, fill = "black", color = "grey") +
  geom_label(stat = "stratum", label.strata = TRUE) +
  scale_x_discrete(limits = c("Gender", "Race"), expand = c(.05, .05)) +
  scale_fill_brewer(type = "qual", palette = "Spectral",name="Legal Action Taken") +
  ggtitle("Was Legal Action Taken? By Driver Gender and Driver Race")
```

\newpage
```{r, warning=F}
# What about just genders?
likelyAlluvia3 <- df %>% 
  mutate(legal_action_taken = Result_of_Stop == "Citation Issued" | Result_of_Stop == "Arrest") %>% 
  group_by(legal_action_taken, Officer_Gender, Driver_Gender) %>% 
  summarise(Frequency=n())

is_alluvia_form(likelyAlluvia3, axes=1:3, silent=TRUE)

likelyAlluvia3 %>% 
  ggplot(aes(y = Frequency, axis1 = Officer_Gender, axis2=Driver_Gender )) +
  geom_alluvium(aes(fill = legal_action_taken), width = 1/12) +
  geom_stratum(width = 1/12, fill = "black", color = "grey") +
  geom_label(stat = "stratum", label.strata = TRUE) +
  scale_x_discrete(limits = c("Officer Gender", "Driver Gender"), expand = c(.1, .1)) +
  scale_fill_brewer(type = "qual", palette = "Set1", name="Legal Action Taken") +
  ggtitle("Was Legal Action Taken? By Officer and Driver Genders")
```



